Predicting the 2024 U.S. Presidential Election: Trump’s Likely Victory Based on Polling Analysis

### 2024-11-03 paper/paper.qmd
committing changes to model and predictions in paper.qmd

library(tidyverse)
library(janitor)
library(lubridate)
library(modelsummary)
library(rstanarm)
library(usmap)
library(arrow)
library(knitr)
library(Matrix)
library(splines)
library(kableExtra)
library(broom.mixed)
```

@@

## Variables
To give an overview of the data, we provide summary statistics of the key variables, highlighting the characteristics of the polling support for each candidate across the battleground states.
```{r}
#| label: fig-summarystatistics
#| fig-cap: "Vertical Summary Statistics for Trump and Harris Polling Data. This table presents key summary statistics, including total polls, average support percentages, and the maximum and minimum polling levels for both candidates across the specified battleground states. All values are rounded to three decimal places, providing a concise overview of the polling landscape as of the selected election date."
#| echo: false
#| warning: false
#| message: false

@@

```{r}
#| label: tbl-outcomevar
#| fig-cap: "Average polling support percentages for Donald Trump and Kamala Harris across key battleground states. The table presents the average support levels, indicating Trump’s and Harris’s polling figures in states such as Wisconsin, North Carolina, Georgia, Arizona, Pennsylvania, Michigan, and Nevada. The comparative analysis highlights the varying levels of candidate support in these critical regions, providing insight into the electoral landscape ahead of the 2024 U.S. Presidential Election."
#| echo: false
#| warning: false
#| message: false

@@

## State Level Support Trends Over Time
```{r}
#| echo: false
#| eval: true
#| label: fig-state-polling-average
#| fig-cap: "Polling Averages for Trump and Harris Over Time in Battleground States."
#| warning: false

@@

## Probability of Winning by State on Election Day

```{r}
#| echo: false
#| eval: true
#| label: tbl-state-probabilities
#| fig-cap: "Probability of Trump Winning by State on Election Day."
#| warning: false

# Set election day or target prediction date
election_day <- as.Date("2024-11-05")  # Example election day

# Predict probability of Trump and Harris winning by comparing posterior predictions for each state
state_predictions <- clean_poll_data %>%
  distinct(state) %>%
  filter(state %in% battleground_states) %>%  # Only include battleground states
  mutate(end_date = election_day, 
         end_date_num = as.numeric(end_date - min(clean_poll_data$end_date)))

# Calculate probabilities of Trump and Harris winning in each state
state_predictions <- state_predictions %>%
  rowwise() %>%
  mutate(
    prob_trump_wins = mean(posterior_predict(trump_model, newdata = cur_data()) > 
                           posterior_predict(harris_model, newdata = cur_data())),
    prob_harris_wins = mean(posterior_predict(harris_model, newdata = cur_data()) > 
                            posterior_predict(trump_model, newdata = cur_data()))
  )

# Prepare the state predictions for the table
state_summary_table <- state_predictions %>%
  select(state, prob_trump_wins, prob_harris_wins) %>%
  mutate(
    prob_trump_wins = round(prob_trump_wins * 100, 2),  # Convert to percentage and round
    prob_harris_wins = round(prob_harris_wins * 100, 2)  # Convert to percentage and round
  )

# Display state predictions in table
knitr::kable(state_summary_table) %>%
  kable_styling(full_width = FALSE, position = "center")

@@

# Reshape the data for plotting only for battleground states
state_predictions_long <- state_predictions %>%
  select(state, prob_trump_wins, prob_harris_wins) %>%
  filter(state %in% battleground_states) %>%  # Ensure we're only working with battleground states
  pivot_longer(cols = starts_with("prob"), 
               names_to = "Candidate", 
               values_to = "Probability") %>%
  mutate(Candidate = recode(Candidate, 
                            prob_trump_wins = "Trump", 
                            prob_harris_wins = "Harris"))

# Plot double bar graph of predicted winning percentages by state
ggplot(state_predictions_long, aes(x = reorder(state, Probability), y = Probability * 100, fill = Candidate)) +
  geom_col(position = "dodge") +  # Use dodge for side-by-side bars
  coord_flip() +  # Flip coordinates for better readability
  scale_fill_manual(values = c("Trump" = "red", "Harris" = "blue")) +
  labs(title = "Predicted Winning Percentages by Battleground State on Election Day", 
       x = "State", 
       y = "Winning Probability (%)") +
  theme_minimal() +
  theme(legend.position = "right")

@@

# Add predicted winner column based on `prob_trump_wins`
state_predictions <- state_predictions %>%
  filter(state %in% battleground_states) %>%  # Ensure we only include battleground states
  mutate(predicted_winner = ifelse(prob_trump_wins > 0.5, "Trump", "Harris"))

# Plot the US map with predicted winner for each state
plot_usmap(data = state_predictions, values = "predicted_winner", regions = "states") +
  scale_fill_manual(values = c("Trump" = "red", "Harris" = "blue")) +
  labs(title = "Predicted Winner by Battleground State on Election Day", fill = "Predicted Winner") +
  theme(legend.position = "right")

>>>>

library(tidyverse)
library(janitor)
library(lubridate)
library(modelsummary)
library(rstanarm)
library(usmap)
library(arrow)
library(knitr)
library(Matrix)
library(splines)
library(kableExtra)
library(broom.mixed)
library(brms)
```

@@

## Variables
To give an overview of the data, we provide summary statistics of the key variables, highlighting the characteristics of the polling support for each candidate across the battleground states.
```{r}
#| label: fig-summarystatistics
#| fig-cap: "Summary Statistics for Trump and Harris Polling Data. This table presents key summary statistics, including total polls, average support percentages, and the maximum and minimum polling levels for both candidates across the specified battleground states. All values are rounded to three decimal places, providing an overview of the polling landscape as of the selected election date."
#| echo: false
#| warning: false
#| message: false

@@

```{r}
#| label: tbl-outcomevar
#| fig-cap: "Summary of fixed effects estimates from the Bayesian hierarchical models for both Donald Trump and Kamala Harris. The table includes the estimated coefficients, standard errors, and 95% confidence intervals for each predictor, illustrating how various factors influence polling percentages. The significant intercept and time-related variables are crucial for understanding candidate support dynamics over the election period."

#| echo: false
#| warning: false
#| message: false

@@

## Polling Averages for Trump and Harris Over Time in Battleground States
```{r}
#| echo: false
#| eval: true
#| label: fig-state-polling-average
#| fig-cap: "Polling Averages for Trump and Harris Over Time in Battleground States."
#| warning: false

@@

## Probability of Winning by State on Election Day 
```{r}
#| echo: false
#| eval: true
#| label: tbl-state-probabilities
#| fig-cap: "Probability of Trump Winning by State on Election Day."
#| warning: false

# Step 1: Prepare new data frame for November 5, 2024
election_day <- as.Date("2024-11-05")
min_date <- min(clean_poll_data$end_date, na.rm = TRUE)
end_date_num_election <- as.numeric(election_day - min_date)

# Create a data frame for predictions, one row per state
new_data <- data.frame(
  state = unique(clean_poll_data$state),
  end_date_num = rep(end_date_num_election, length(unique(clean_poll_data$state)))
)

# Step 2: Generate predictions for both models
trump_predictions <- posterior_predict(trump_model, newdata = new_data)
harris_predictions <- posterior_predict(harris_model, newdata = new_data)

# Step 3: Calculate winning percentages
# Calculate mean predictions and convert to percentages
trump_mean_pct <- apply(trump_predictions, 2, mean) * 100
harris_mean_pct <- apply(harris_predictions, 2, mean) * 100

# Combine results into a data frame
state_predictions <- data.frame(
  state = new_data$state,
  trump_pct = trump_mean_pct,
  harris_pct = harris_mean_pct
)

# Round the percentages to two decimal points
state_predictions <- state_predictions %>%
  mutate(
    trump_pct = round(trump_pct, 2),
    harris_pct = round(harris_pct, 2),
    winner = ifelse(trump_pct > harris_pct, "Donald Trump", "Kamala Harris")
  )

# Print results in table
kable(state_predictions, 
      col.names = c("State", 
                    "Predicted Harris Win (%)", 
                    "Predicted Trump Win (%)", 
                    "Predicted Winner")) %>%
  kable_styling("striped", full_width = F)

@@

# Reshape the data for ggplot
state_predictions_long <- predicted_results %>%
  pivot_longer(cols = c(harris_pct, trump_pct), 
               names_to = "candidate", 
               values_to = "winning_probability") %>%
  mutate(candidate = recode(candidate, 
                            harris_pct = "Kamala Harris", 
                            trump_pct = "Donald Trump"))

# Create the bar chart
ggplot(state_predictions_long, aes(x = state, y = winning_probability, fill = candidate)) +
  geom_bar(stat = "identity", position = "dodge") +
  coord_flip() +  # Flip coordinates for horizontal bars
  scale_fill_manual(values = c("blue", "red")) +  # Specify colors
  labs(title = "Predicted Winning Percentages by Battleground State on Election Day",
       x = "Winning Probability (%)",
       y = "State") +
  theme_minimal() +
  theme(legend.title = element_blank())  # Remove legend title

@@

# Step 4: Plot the US map with predicted winner for each state
# Convert state names to lowercase to match the map data
state_predictions$state <- tolower(state_predictions$state)

# Create the map
us_map <- plot_usmap(data = state_predictions, values = "winner", regions = "states") +
  scale_fill_manual(values = c("Donald Trump" = "red", "Kamala Harris" = "blue", "NA" = "gray"), 
                    na.value = "gray") +
  labs(title = "Predicted Winner by Battleground State on Election Day", 
       fill = "Predicted Winner") +
  theme(legend.position = "right")

# Print the map
print(us_map)
```
